---
title: "Intro to Data Frames in R"
subtitle: "Environmental Data Analytics"
author: "Luana Lima"
date: "Fall 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data frame definition

A **data frame** is used for storing data tables. It a list a vector of equal length.

When we import data to R, data frame is the preferred way for storing the data because columns can have different modes (character, numeric, integer, logical, complex).

## Data frame built-in example

Let's look into a built-in data frame from package "datasets" - cars. The data give the speed of cars and the distances taken to stop.

```{r cars}
cars
mtcars
```

Note that it has 2 columns and 50 rows.

## Data frame columns

Suppose you want just the column speed. How would you access that data?

```{r}
cars$speed
mtcars$cyl
```

How would you store it on another object?

```{r}
car_speed <- cars$speed
car_speed
```

## Transforming object in a data frame

Is the new object you create a data frame?

```{r}
#Option 1
class(car_speed)

#Option 2
is.data.frame(car_speed)
is.vector(car_speed)
```

How could you make it a data frame?

```{r}
df_car_speed <- as.data.frame(car_speed)
df_car_speed
class(df_car_speed)
```

## Adding columns to a data frame

How could you add columns to *df_car_speed*?

```{r}
#cbind can make multiple vetors a data frame, or multiple data frames into a larger data frame

car_dist <- cars$dist

#Option 1
df <- cbind(df_car_speed,car_dist)  #similarly rows could be added using rbind()
class(df)
df

#Or Option 2 - transform into a data frame before binding
df_car_dist <- as.data.frame(car_dist)  #op2
df_opt2 <- cbind(df_car_speed,df_car_dist)
class(df_opt2)
df_opt2

```

Note that when we transformed the vector in a data frame the name of the vector became the column name.

```{r}
colnames(df)  #or simply names()
names(df)
```

## Creating a data frame

How would you create a data frame?

```{r}
#useful function data.frame()
#creating data frame from two vectors, giving names to each vector
create_df <- data.frame("speed"=car_speed,"dist"=car_dist)
create_df
```

## Data frame functions

Some useful functions to use with data frames.

```{r}
ncol(df)  #number of columns
nrow(df)  #number of rows
length(df)  #same as ncol
summary(df) #quick summary stats
head(df) #show the first 6 rows of df
head(df,2) #customize how many rows are shown

#If you know the number of the column you want you can refer to that to access column

df2 <- as.data.frame(df$car_speed) [c(2,3,1),]
df2

df[,1]
df[1:10,2] #gives the first through tenth values of column two
df[1:10,] #gives the first through tenth values of all columns
df[c(2,3,1),1] #gives values 2,3,1 of column one (customized order)
df[c(2,3,1),] #gives value 2,3,1 of all columns (customized order)

#you could also use this notation to delete columns
df <- df[,-2]
df
```
