---
title: "Assignment 10: Data Scraping"
author: "Hannah Nelson"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(here)
library(rvest)
library(tidyverse)
library(tidycensus)
library(dataRetrieval)
library(lubridate)

here()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
site <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022")

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 

wsysname <- site %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text

wsysname
PWSID 
ownership
MGD

MGD <- c(36.1000, 30.5000, 39.9100, 34.6600, 43.4200, 42.5900, 43.3200, 
         41.8000, 52.4900, 34.8800, 32.5300, 37.5300)

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 

durham_2022 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

durham_2022$Month <- factor(durham_2022$Month, 
                            levels=c("Jan", "Feb", "Mar", "Apr", "May",
                                     "Jun", "Jul", "Aug", "Sep", "Oct",
                                     "Nov","Dec"))

durham_2022

#5 

ggplot(durham_2022, aes(y = MGD, x = Month, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Maximum Daily Withdrawals for Each Month of 2022", 
       subtitle = "Durham, NC")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
scrape <- function(PWSID, year) {
  url <- paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", 
                  PWSID, "&year=", year)
  
  page <- read_html(url)
  
  wsysname <- site %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text
  
  PWSID <- site %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text
  
  ownership <- site %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text
  
  MGD <- site %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text
  
  year <- 2022
  
  months <- month.abb

}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 

site2 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2015")

wsysname <- site2 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text

wsysname
PWSID 
ownership
MGD


durham_2015 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

durham_2015$Month <- factor(durham_2015$Month, 
                            levels = c("Jan", "Feb", "Mar", "Apr", "May", 
                                       "Jun", "Jul", "Aug", "Sep", "Oct", 
                                       "Nov", "Dec"))

durham_2015


ggplot(durham_2015, aes(y = MGD, x = Month, group = 1)) +
  geom_point() +
  geom_line() +
  labs(title = "Maximum Daily Withdrawals for Each Month of 2015", 
       subtitle = "Durham, NC")


```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
site3 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2015")

wsysname <- site3 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site3 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site3 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site3 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text

wsysname
PWSID 
ownership
MGD


asheville_2015 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2015$Month <- factor(asheville_2015$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))

asheville_2015



ggplot() +
  geom_line(data = asheville_2015, 
            aes(x = Month, y = MGD, group = 1, color = Water.System.Name)) +
  geom_point(data = asheville_2015, 
            aes(x = Month, y = MGD, group = 1, color = Water.System.Name)) +
  geom_line(data = durham_2015, 
            aes(x = Month, y = MGD, group = 1, color = Water.System.Name)) +
  geom_point(data = durham_2015, 
            aes(x = Month, y = MGD, group = 1,  color = Water.System.Name)) +
    labs(title = "Maximum Daily Withdrawals for Each Month of 2015", 
       subtitle = "For Durham & Asheville, NC", 
       color = "Water System Name")
                     


```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 

site2010 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2010")

wsysname <- site2010 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2010 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2010 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2010 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text


asheville_2010 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2010$Month <- factor(asheville_2010$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))


site2011 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2011")

wsysname <- site2011 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2011 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2011 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2011 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text


asheville_2011 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2011$Month <- factor(asheville_2011$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))



site2012 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2012")

wsysname <- site2012 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2012 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2012 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2012 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text


asheville_2012 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2012$Month <- factor(asheville_2012$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))

site2013 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2013")

wsysname <- site2013 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2013 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2013 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2013 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text

asheville_2013 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2013$Month <- factor(asheville_2013$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))

site2014 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2014")

wsysname <- site2014 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2014 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2014 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2014 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text

asheville_2014 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2014$Month <- factor(asheville_2014$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))


site2015 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2015")

wsysname <- site2015 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2015 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2015 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2015 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text


asheville_2015 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2015$Month <- factor(asheville_2015$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))

site2016 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2016")

wsysname <- site2016 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2016 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2016 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2016 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text


asheville_2016 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2016$Month <- factor(asheville_2016$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))


site2017 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2017")

wsysname <- site2017 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2017 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2017 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2017 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text


asheville_2017 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2017$Month <- factor(asheville_2017$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))



site2018 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2018")

wsysname <- site2018 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2018 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2018 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2018 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text


asheville_2018 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2018$Month <- factor(asheville_2018$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))


site2019 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2019")

wsysname <- site2019 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2019 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2019 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2019 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text


asheville_2019 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2019$Month <- factor(asheville_2019$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))



site2020<-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2020")

wsysname <- site2020 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2020 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2020 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2020 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text


asheville_2020 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2020$Month <- factor(asheville_2020$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))



site2021 <-read_html(
  "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=01-11-010&year=2021")

wsysname <- site2021 %>%
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>%
  html_text

PWSID <- site2021 %>%
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>%
  html_text

ownership <- site2021 %>%
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>%
  html_text


MGD <- site2021 %>%
  html_nodes("th~ td+ td , th~ td+ td") %>%
  html_text


asheville_2021 <- data.frame(stringsAsFactors = T,
                 "Water System Name" = wsysname,
                 "Ownership" = ownership,
                 "PWSID" = PWSID,
                 "Month" = month.abb,
                 "MGD" = MGD)

asheville_2021$Month <- factor(asheville_2021$Month, 
                               levels=c("Jan","Feb","Mar","Apr","May","Jun",
                                      "Jul","Aug","Sep","Oct","Nov","Dec"))



ggplot() +
  geom_line(data = asheville_2010, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2011, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2012, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2013, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2014, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2015, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2016, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2017, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2018, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2019, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2020, 
            aes(x = Month, y = MGD, group = 1)) +
  geom_line(data = asheville_2021, 
            aes(x = Month, y = MGD, group = 1))
  
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
>
