---
title: "Assignment 2: Coding Basics"
author: "Hannah Nelson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
#creating a sequence from 1-30 that counts by 3
sequence <- seq(1, 30, by=3)
sequence

#2. 
#finding the mean and median values of the sequence and naming those objects
mean <- mean(sequence)
mean
#the mean is 14.5
median <- median(sequence)
median
#the median is 14.5

#3. 
#creating a function that will indicate if the mean is larger than the median by generating true/false
measures <- function(sequence) {
  measures <- ifelse(mean > median, "TRUE", "FALSE")
  print(measures)
}

#creating an if else function to generate text indicating if the mean is greater, less than, or equal to the median 
#this is useful because I know the mean and median are the same value, but that could not be reflected using the true/false function above
measures2 <- function(sequence) {
  measures2 <- if (mean > median) {
    print("The mean is greater than the median")
  }
  else if (mean < median) {
  print("The mean is less than the median")
  }
  else {
  print("The mean and median are equal")
  }
}


#run sequence through functions to generate results
#this function tells us that the mean is not larger than the median
measures(sequence)
#this function tells us the mean is equal to the median
measures2(sequence)
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
#5. & #6.
#creating vector with four names called "names"
names <- c("Bella", "Emily", "Alex", "Rachel")
#creating vector with four scores called "scores"
scores <- c(80, 90, 70, 40)
#creating vector with true/false values indicating if a passing score was achieved corresponding to the values in the scores vector called "pass"
pass <- c("TRUE", "TRUE", "TRUE", "FALSE")

#7. & #8.
#creating data frame called "testing" by combining the three vectors and naming the columns
testing <- data.frame("names"=names, "scores"=scores, "passing_score"=pass)
testing
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: This data frame contains multiple types of data, while a matrix can only contain one type.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#scores vetor from question 5:
scores <- c(80, 90, 70, 40)

#10.
#function that prints "true" if score is 50 or higher or "false" if it is lower than 50
passfail <- function(scores) {
  passfail <- ifelse(scores >= 50, "TRUE", "FALSE")
  print(passfail)
}

#11.
#run function on vector to produce true/false list indicating if each score in the vector was passing
passfail(scores)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: the "ifelse" function worked becuase it allows you to apply a function to an entire vector, while "if" and "else" do not.
